---
title: "Debugging R Code"
---

Debugging is an essential skill for any R programmer. It's the process of 
finding and fixing errors or unexpected behaviors in your code. This chapter
will cover various techniques and tools to help you effectively debug R code.

## Understanding Error Messages

The first step in debugging is to understand the error messages R provides.
R error messages typically include:

- The type of error
- The line number where the error occurred
- A brief description of the problem

For example:

```r
Error in mean(x) : object 'x' not found
```

This error tells us that the function `mean()` was called with an object `x`,
but `x` doesn't exist in the current environment. From this message, we can 
infer that `x` is not defined or is out of scope.

## Using `print()` Statements

One of the simplest debugging techniques is to use `print()` statements 
throughout your code. This helps you track the values of variables and the 
flow of your program.

```{r}
calculate_mean <- function(data) {
  print("Entering function")
  print(paste("Data:", data))
  result <- mean(data)
  print(paste("Result:", result))
  return(result)
}
```

When you run this function, you'll see the printed messages in the console,
which can help you identify where things are going wrong.

```{r}
calculate_mean(c(1, 2, 3, 4, 5))
```

### Using Diagnostic Messages

In addition to `print()` statements, you can use `message()`, `warning()`, and
`stop()` to provide more informative diagnostic messages.

#### `message()`

The `message()` function is similar to `print()` but is intended for 
informational messages rather than debugging output.

```{r}
calculate_mean <- function(data) {
  if (length(data) == 0) {
    message("Data is empty")
  }
  result <- mean(data)
  return(result)
}
```

When you run this function with an empty data vector, you'll see the message
"Data is empty" in the console. The function will continue to execute, but the
message can help you identify potential issues.

#### `warning()`

The `warning()` function is used to display non-fatal issues that may affect
the results of your code.

```{r}
calculate_mean_with_warning <- function(data) {
  if (length(data) == 0) {
    warning("Data is empty")
  }
  result <- mean(data)
  return(result)
}
```

When you run this function with an empty data vector, you'll see a warning
message in the console. The function will continue to execute, but the warning
can help you identify potential problems. We can also use `warnings()` to
retrieve a list of all warnings that have been generated.

```{r}
warnings()
```

We can modify the behavior of warnings by setting the global option `warn`. 
In particular, `warn` offers three levels of support:

1. `options(warn = 0)` will suppress all warnings until the top-level function has completed.
2. `options(warn = 1)` will display warnings as they occur but not halt execution.
3. `options(warn = 2)` converts warnings into errors that will halt execution.

For debugging purposes, you may wish to use `options(warn = 2)` to catch
issues early in your code. 

#### `stop()`

The `stop()` function allows you to halt execution and display an error message.
This can be useful for catching unexpected conditions in your code.

```{r}
#| error: true
calculate_mean_with_stop <- function(data) {
  if (length(data) == 0) {
    stop("Data is empty")
  }
  result <- mean(data)
  return(result)
}

calculate_mean_with_stop(numeric())
```


## Using `browser()`

The `browser()` function is a powerful tool for interactive debugging. When R 
encounters a `browser()` statement, it pauses execution and allows you to
inspect the current environment.

```{r}
calculate_mean_browser <- function(data) {
  browser()
  result <- mean(data)
  return(result)
}

calculate_mean_browser(1:5)
```

When you run this function, R will pause at the `browser()` call, allowing you 
to examine and manipulate variables. This is particularly useful for 
understanding the state of your program at a specific point in time. 

When you're in the browser, you can navigate through the code using the
following commands:

- `ls()` to list the objects in the current environment.
- `print(object)` to display the value of an object.
- `c` to continue execution by exiting the browser.
- `f` to finish the current loop or function.
- `n` to step to the next line.
- `s` to step into a function call.
- `where` to display the call stack.
- `r` to restart if you want to re-run the function.
- `Q` to quit the debugger.

We'll see in @sec-rstudio-debugging-tools that RStudio provides a more 
user-friendly interface for this type of debugging through icons and breakpoints.

## Using an interactive debugger

The `debug()` function allows you to step through a function line by line. We
can use `debug()` to immediately start debugging a function when it is 
called. For example:

```{r}
debug(calculate_mean)
calculate_mean(c(1, 2, 3, 4, 5))
```

Notice, the function `calculate_mean` is now in debug mode. When you run the
function, R will pause at the first line of the function and allow you to step
through the code. You can use similar commands as we saw with `browser()` to
inspect variables and control the flow of the program.

If we want to remove the debugger, we would use:

```{r}
undebug(calculate_mean)
```

Though, if we only want to debug the next call, we can use `debugonce()` instead.
This will only debug the next call to the function without needing to remove the
debugger afterwards.


## Using `traceback()`

`traceback()` shows you the sequence of function calls that led to an error. It's particularly useful for understanding errors in complex, nested function calls.

```{r}
#| error: true
f <- function(x) g(x)
g <- function(x) h(x)
h <- function(x) x + "a"

f(10)  # This will cause an error
```

```r
traceback()
# 3. h(x)
# 2. g(x)
# 1. f(10)
```

## Implementing Error Handling

TODO: Add a section describing fail-fast principles

Proper error handling can make debugging easier. Use `try()`, `tryCatch()`, and custom error messages to manage potential issues.

```{r}
safe_mean <- function(x) {
  tryCatch(
    mean(x),
    error = function(e) {
      message("An error occurred: ", e$message)
      return(NA)
    }
  )
}

safe_mean("bad input")
```

In this example, `safe_mean()` attempts to calculate the mean of a vector. 
If an error occurs, it prints a message and returns `NA` instead of halting 
execution.

## Using RStudio's Debugging Tools  {#sec-rstudio-debugging-tools}

If you're using RStudio, take advantage of its built-in debugging tools:

- Breakpoints: Click next to a line number to set a breakpoint.
- Environment pane: Inspect variable values during debugging.
- Debug toolbar: Step through code, examine the call stack, and more.

TODO: Add screenshot of the debugger

## Leveraging Package-Specific Debugging Tools

Some R packages provide their own debugging tools. For example, the `debugr`
package offers a more streamlined print debugging experience.

TODO: Add an example of using debugr


## Common Debugging Pitfalls

Be aware of common issues that can complicate debugging:

- Scope issues: Ensure you're looking at the correct environment.
- Data type mismatches: Check that your functions are receiving the expected data types.
- Missing values: Be cautious of `NA` values affecting your calculations.

## Best Practices for Debuggable Code

Write code that's easier to debug:

- Use meaningful variable names
- Break complex operations into smaller, testable functions
- Comment your code thoroughly
- Use version control to track changes


## Debugging in Quarto 

TODO: Add a section on debugging with Quarto

## Summary

Through mastering these debugging techniques and tools, you'll be well-equipped to
tackle even the most challenging coding issues in R. Remember, effective 
debugging is not just about fixing errors—it's about understanding your code 
more deeply and writing more robust programs.